### 2025-10-02 thesis/design/definitions.tex
Improved the definition and sampling sections

\begin{enumerate}
    \item $\mathcal{Q}$ – The set of all possible natural language queries
    \item $\mathcal{S}$ – The set of all possible database schemas
    \item $\mathcal{E}$ – The set of execution functions for SQL validation
    \item $\mathcal{V}$ – The embedding space of historically observed samples
    \item $\mathcal{D}$ – The space of distance indices over historically observed databases
    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
    \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
\end{enumerate}

@@

where $q \in \mathcal{Q}$ is the input natural language query, $z \in \Omega_{\mathcal{S}}$ is the zero-shot inference result, $s \in \mathcal{S}$ is the database schema, $v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index, and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$ tuples of natural language queries, schemas, corresponding SQL queries and their combined similarity distances.

@@

\begin{enumerate}
    \item \textbf{Question similarity} – Semantic similarity between the masked input query and historically observed natural language queries in the embedding space $\mathcal{V}$, measured using cosine distance between their respective embeddings
    \item \textbf{Query similarity} – Structural similarity between the zero-shot inference result $z$ and candidate SQL queries using masked SQL representations, enabling pattern recognition across different database domains
    \item \textbf{Database similarity} – Schema compatibility measured through the distance index $\mathcal{D}$, ensuring selected examples operate on analogous database structures with similar table relationships and column types
\end{enumerate}

@@

The subsetting function $\phi$ intelligently reduces the database schema to only include
tables, columns, and relationships that are relevant to the current query context.
This schema pruning mechanism reduces the complexity of the SQL generation task and
improves model performance by focusing on pertinent schema elements.

@@

The function analyzes the selected examples in $c$ to identify which schema elements
(tables, columns, foreign key relationships) are commonly referenced in similar queries.
This analysis enables the system to steer the model's attention on the most relevant
portions of potentially large and complex database schemas, thereby improving both accuracy
and computational efficiency.

@@

The projection function $\pi$ represents the core translation mechanism that converts
natural language queries into SQL statements using the selected examples and reduced schema.
This function encapsulates the language model inference process that generates candidate
SQL queries based on the provided context.

$$
\pi: \mathcal{Q} \times \mathcal{S}' \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}'}
$$

@@

$$
\pi(q, s', c) = \omega
$$

@@

where $q \in \mathcal{Q}$ is the input natural language query, $s' \in \mathcal{S}'$ is
the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
and $\omega \in \Omega_{\mathcal{S}'}$ is the generated SQL query.

@@

$$
\rho: \mathcal{Q} \times \mathcal{S}' \times \mathcal{E} \times \Omega_{\mathcal{S}'} \rightarrow \Omega_{\mathcal{S}'}
$$

@@

$$
\rho(q, s', e, \omega_{prev}) = \omega_{refined}
$$

@@

where $q \in \mathcal{Q}$ is the original natural language query, $s' \in \mathcal{S}'$
is the relevant schema subset, $e \in \mathcal{E}$ is the execution environment for
validation, $\omega_{prev} \in \Omega_{\mathcal{S}'}$ is the previously generated SQL
query, and $\omega_{refined} \in \Omega_{\mathcal{S}'}$ is the improved SQL query.

@@

The voting function $\nu$ implements a consensus mechanism that selects the most
reliable SQL query from multiple candidate solutions generated through the pipeline.
This function enhances robustness by leveraging the wisdom of multiple generation
attempts to identify the most likely correct answer.

$$
\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}'}
$$

@@

where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
the execution environment for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}'}$
is the selected consensus query.

@@

The voting function $\nu$ implements a majority voting algorithm similar to that
described by OmniSQL \citep{OmniSQL}, where the result that appears most frequently
across multiple generation attempts is deemed to be the most likely correct answer.
The function first validates all candidates for syntactic correctness and executability,
then applies frequency-based selection among the valid candidates. In cases where
no clear majority exists, the function may apply additional heuristics such as query
complexity or execution performance to determine the final selection.

>>>>

\begin{enumerate}
    \item $\mathcal{Q}$ – The set of all possible natural language queries
    \item $\mathcal{S}$ – The set of all possible database schemas
    \item $\mathcal{E}$ – The set of all possible execution functions for SQL validation
    \item $\mathcal{V}$ – The set of all possible embedding spaces
    \item $\mathcal{D}$ – The set of all possible distance indices over historically observed databases
    \item $\mathcal{C}$ – The set of all possible candidate sets $\mathcal{C}'$
    \item $\mathcal{Q}_{\mathcal{S}}$ – The set of all valid queries over the database schema $S$.
\end{enumerate}

@@

where $q \in \mathcal{Q}$ is the input natural language query, $z \in \Omega_{\mathcal{S}}$
is the zero-shot inference result, $s \in \mathcal{S}$ is the database schema,
$v \in \mathcal{V}$ is the embedding space, $d \in \mathcal{D}$ is the distance index,
and the result is a candidate set $\mathcal{C}' \in \mathcal{C}$ containing $k$
tuples of natural language queries, schemas, corresponding SQL queries and their
combined distance.

@@

\begin{enumerate}
    \item \textbf{Question similarity} – Semantic similarity between the masked input
        query and historically observed natural language queries in the embedding space
        $\mathcal{V}$, measured using cosine distance between their respective embeddings.
    \item \textbf{SQL similarity} – Structural similarity between the zero-shot inference
        result $z$ and candidate SQL queries using masked SQL representations, enabling
        pattern recognition across different database domains.
    \item \textbf{Database similarity} – Schema compatibility measured through the distance
        index $\mathcal{D}$, ensuring selected examples operate on analogous database
        structures with similar table relationships and column types.
\end{enumerate}

The exact weighting of these can be adjusted through three constants, $w_q}$,
$w_s}$, and $w_d}$.

@@

The subsetting function $\phi$ reduces the database schema to only include
tables, columns, and relationships that are relevant to the current set of candidates
$C$. This schema pruning mechanism reduces the complexity of the SQL generation task
and token efficiency by focusing on the most relevant schema elements.

@@

The function analyzes the selected examples in $c$ to identify which schema elements
(tables, columns, foreign key relationships) are commonly referenced in similar queries.
This analysis enables the system to steer the model's attention on the most relevant
portions of potentially large and complex database schemas, thereby improving both accuracy
and computational efficiency and prevents exceeding limited context windows.

@@

The projection function $\pi$ represents the core translation mechanism that converts
natural language queries into SQL statements using the selected examples and database
schema. This function encapsulates the large language model inference that generates
candidate SQL queries based on the provided context.

$$
\pi: \mathcal{Q} \times \mathcal{S} \times \mathcal{C}' \rightarrow \Omega_{\mathcal{S}}
$$

@@

$$
\pi(q, s, c) = \omega
$$

@@

where $q \in \mathcal{Q}$ is the input natural language query, $s \in \mathcal{S}$ is
the reduced schema subset, $c \in \mathcal{C}'$ is the set of selected examples,
and $\omega \in \Omega_{\mathcal{S}}$ is the generated SQL query.

@@

$$
\rho: \mathcal{Q} \times \mathcal{S} \times \mathcal{E} \times \Omega_{\mathcal{S}} \rightarrow \Omega_{\mathcal{S}}
$$

@@

$$
\rho(q, s, e, \omega_{raw}) = \omega_{refined}
$$

@@

where $q \in \mathcal{Q}$ is the original natural language query, $s \in \mathcal{S}$
is the database schema, $e \in \mathcal{E}$ is the execution function for
validation, $\omega_{raw} \in \Omega_{\mathcal{S}}$ is the previously generated SQL
query and $\omega_{refined} \in \Omega_{\mathcal{S}}$ is the improved SQL query.

@@

The voting function $\nu$ implements a consensus mechanism that selects the most
reliable SQL query from multiple candidate solutions generated through the pipeline.
This function enhances robustness by leveraging the result distribution of multiple
generation attempts to identify the most likely correct answer.

$$
\nu: \mathcal{C} \times \mathcal{E} \rightarrow \Omega_{\mathcal{S}}
$$

@@

where $C \in \mathcal{C}$ is a set of candidate SQL queries, $e \in \mathcal{E}$ is
the execution function for validation, and $\omega_{consensus} \in \Omega_{\mathcal{S}}$
is the selected consensus query.

@@

The voting function $\nu$ implements a majority voting algorithm similar to that
described by OmniSQL \citep{OmniSQL}, where the result that appears most frequently
across multiple generation attempts is deemed to be the most likely correct answer.
The function applies frequency-based selection among the valid candidates. In cases where
no clear majority exists, the function may apply additional heuristics such as query
complexity or execution performance to determine the final proposed query candidate $w_{consensus}$.

### 2025-10-02 thesis/design/index.tex
Improved the definition and sampling sections

\caption{The $nq$ function}\label{alg:cap}

@@

Prior to using $nq$ sample and database indexing must be done to supply a relevant 
embedding space $\mathcal{V}$ and distance index $\mathcal{D}$.

\subimport{}{sampling}
\subimport{}{definitions}

>>>>

\caption{The $nq$ function}\label{algorithms:nq}

@@

Prior to using $nq$ to answer natural language questions, historical datasets must be sampled
such that an embedding space $\mathcal{V}$ and distance index $\mathcal{D}$ can be supplied
to $nq$.

\subimport{}{sampling}

\subimport{}{definitions}

### 2025-10-02 thesis/design/sampling.tex
Improved the definition and sampling sections

%Masked Embedding Space – $\mathcal{V}$

@@

The natural language masking process replaces all words which are not deemed
\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
Choosing a sound set $\mathcal{W}$ is described in chapter 5.

This process enables the embedding space to capture semantic patterns independent
of specific database artifacts and domain terminology, improving the generalizability
of example selection across different schemas.

This abstraction allows the system to recognize that queries like:

@@

in fact share similar semantic structures, even though they operate on different entities.
Given $\mathcal{W} = \{\text{List}\, \text{Show}\, \text{all}\, \text{the}\, \text{with}\, \text{name}\}$

@@

Query masking applies similar abstraction principles to the SQL queries themselves.
This enables the system to identify SQL patterns and structures rather than
memorizing specific database schemas.

The SQL masking process transforms all identifiers to \texttt{<mask>} and all literals
to \texttt{<value>}. For example, the query:

@@

This abstraction enables pattern recognition across different domains while
preserving the essential logical structure of the original SQL statements.

>>>>

@@

The natural language masking process replaces all words which are not deemed
\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
% TODO: Fix reference
Choosing a sound set $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.

This process enables the embedding space to capture semantic patterns independently
of specific database artifacts and domain terminology, improving the generalizability
across different schemas and domains.

This masking allows \textsc{Natural} to recognize that queries like:

@@

in fact share similar semantic structures, even though they operate on different entities.
Given $\mathcal{W} = \{\text{List}, \text{Show}, \text{all}, \text{the}, \text{with}, \text{name}\}$

@@

Query masking applies similar abstraction principles to the corresponding SQL queries.
This enables \textsc{Natural} to identify SQL patterns and structures.

The SQL masking process transforms all identifiers to \texttt{<mask>} and all
literals to \texttt{<value>}. Thus, the query:

@@

This abstraction enables pattern recognition across different domains while
preserving the essential logical structure of the original SQL statements,
thus focusing on the kind of query, rather than the specific instance.

\subsubsection{Sample Embedding Space – $\mathcal{V}$}

\subsubsection{Schema Distance Index – $\mathcal{D}$}

The schema distance index is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
where $s$ is the database schema name and $d$ is the distance to the
\textit{current} schema. Thus a schema distance index has to be computed for each database schema
\textsc{Natural} is running on.
